---
title: "Memteorites"
author: "Elio Campitelli"
output: github_document
---

First read the data:

```{r}
library(data.table)
library(ggplot2)
library(magrittr)

meteorites <- fread("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-06-11/meteorites.csv")
```
```{r}
str(meteorites)
```

Ok. We've got some interesting data. I love having coordinates. I don't know exactly what "geolocation" is. It seems to be some sort of alternative coordinates. I wonder about the "class" column, though. How many class are there?

```{r}
uniqueN(meteorites$class)
```


Woah! That a lost of clases! How many meteors of each? (only looking at "valid" objects)

```{r}
meteorites[name_type == "Valid", .N, by = class] %>% 
  .[, class := reorder(class, -N)] %>% 
  .[1:20] %>% 
  ggplot(aes(class, N)) +
  geom_col()
```

VERY unbalanced. It seems that most meteorites are concentrated in those 6 categories. There's a long wikipedia article that I could read if I decided to analyse them. 

I bet mass distribution is also very skewed. 

```{r}
meteorites %>% 
  ggplot(aes(mass)) +
  geom_histogram() +
  scale_x_log10()
```

Yup. There are 131 missing values, also. Very few, so I'm not gonna worry about it just now. Although, how many complete cases do we have?

```{r}
mean(complete.cases(meteorites))
```

83% is not bad, but also not excellent. 

What's the distribution with age?

```{r}
ggplot(meteorites, aes(year)) +
  geom_histogram(binwidth = 1) +
  scale_x_continuous(limits = c(NA, 2019))
```

Not surprisingly, most of them are registered in moder times, but there are also ancient records. But just a tiny amount:

```{r}
nrow(meteorites[year < 1800])
```

I also want to glance at the spatial distribution. 

```{r}
map <- geom_sf(data = rnaturalearth::ne_countries(returnclass = "sf"), 
              inherit.aes = FALSE)

ggplot(meteorites, aes(long, lat)) +
  map +
  geom_point() 
```

.... That one on the far right is nonsense. 

```{r}
meteorites[, long := metR::ConvertLongitude(long, from = 360)]


ggplot(meteorites, aes(long, lat)) +
  map + 
  geom_point()
```

Much better. Althouth the distribution of falling meteorites is probably uniform, there's clearly a strong bias towards populated areas. 

```{r}
(g <- ggplot(meteorites, aes(long, lat)) +
  map + 
  geom_hex() )
```

Wowzer! Why are there so many of them found in ecuatorial Africa? I bet that's due to a concerned research effort and not accidental finds. If we split according to found vs fell meteorites...

```{r}
g +
  facet_wrap(~fall)
```

Yup. the scale is not working here. I'll concentrate on the "fell" meteorites. 

```{r}
g %+% meteorites[fall == "Fell"]
```


```{r}
ggplot(meteorites[is.finite(mass)], aes(long, lat)) +
  map + 
  stat_summary_hex(aes(z = mass), fun = sum, bins = 80) +
  scale_fill_viridis_c()
```

