---
title: "Memteorites"
author: "Elio Campitelli"
output: github_document
---

First read the data:

```{r}
library(data.table)
library(ggplot2)
library(magrittr)
library(hrbrthemes)

meteorites <- fread("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-06-11/meteorites.csv")

name <- knitr::current_input(dir = TRUE)
name <- "/home/elio/Documents/ddm/03 - Meteorites/README.Rmd"
name <- strsplit(name, "/", fixed = TRUE)[[1]]
dir <- name[length(name) - 1]
dir <- stringi::stri_replace(dir, "%20", fixed = " ")

url <- paste0("https://github.com/eliocamp/ddm/tree/master/", dir)

library(spindler)
this_thread <- thread$new(tag = "tw")

this_thread$add_post("I'm a bit late for this week's #tidytuesday, but I wanted to give it a go anyway. Specially because I'm using it as an excuse to try the spindler package in the real world. So join me as we learn a (little) bit about meteorites! #rstats")$
  add_post("So this week's data consists of about 50000 records of meteorites found on Earth. They've got names, mass, a classification, and long/lat coordinates, all divided by year. ")


Compass <- function(x, y, a , b = a, n = 10) {
  tita <- seq(0, 2*pi, length.out = n + 1)
  x <- x + a*cos(tita)
  y <- y + b*sin(tita)
  data.frame(x = x, y = y )
}

```

```{r}
str(meteorites)
```

Ok. We've got some interesting data. I love having coordinates. I don't know exactly what "geolocation" is. It seems to be some sort of alternative coordinates. I wonder about the "class" column, though. How many class are there?

```{r}
uniqueN(meteorites$class)
```


Woah! That a lost of clases! How many meteors of each? (only looking at "valid" objects)

```{r, tw = "My first though was to see what those 'classes' were. It turn out that even though there are more than 450 unique classes, the vast majority of meteorites belong to just 6 of them"}
meteorites[name_type == "Valid", .N, by = class] %>% 
  .[, class := reorder(class, -N)] %>% 
  .[1:20] %>% 
  ggplot(aes(class, N)) +
  geom_col()
```

VERY unbalanced. It seems that most meteorites are concentrated in those 6 categories. There's a long wikipedia article that I could read if I decided to analyse them. 

```{r}
this_thread$add_post("There's a long-ass wikipedia article on the subject (https://en.m.wikipedia.org/wiki/Meteorite_classification ), but as someone with zero geology training it's all meaningless to me. Lesson one: data withouth appropiate domain expertise is pretty much useless.")
```

```{r}
meteorites[name_type == "Valid", .N, by = class] %>% 
  .[order(-N)] %>% 
  .[1:6] %>% 
  .[, unique(class)] -> most_classes
```


I bet mass distribution is also very skewed. 

```{r, tw = "Another thing I wanted to look at was the mass distribution. As I expected, it's extremely skewed (see the log x axis). There are also (very few) missing values. "}
meteorites %>% 
  ggplot(aes(mass)) +
  geom_histogram() +
  scale_x_log10()
```

Yup. There are 131 missing values, also. Very few, so I'm not gonna worry about it just now. Although, how many complete cases do we have?

```{r}
mean(complete.cases(meteorites))
```

83% is not bad, but also not excellent. 

What's the distribution with age?

```{r}
ggplot(meteorites, aes(year)) +
  geom_histogram(binwidth = 1) +
  scale_x_continuous(limits = c(NA, 2019))
```

Not surprisingly, most of them are registered in moder times, but there are also ancient records. But just a tiny amount:

```{r}
nrow(meteorites[year < 1800])
```

I also want to glance at the spatial distribution. 

```{r}
map <- geom_sf(data = rnaturalearth::ne_countries(returnclass = "sf"), 
               inherit.aes = FALSE, fill = NA, color = "gray", size = 0.2)

ggplot(meteorites, aes(long, lat)) +
  map +
  geom_point() 
```

.... That one on the far right is nonsense. 

```{r, tw = "But what I'm most excited is the coordiantes of the find. It turns out that meteorites don't fall uniformly on earth? No! Of course they do, but people find meteorites and they are not uniformly distributed."}
meteorites[, long := metR::ConvertLongitude(long, from = 360)]


(g <- ggplot(meteorites, aes(long, lat)) +
  map + 
  geom_point(alpha = 0.3, size = 0.2) +
    theme_ipsum_rc())
```

```{r, tw = "But it's not only population density. Look at Antarctica! Finding a meteorite in a sea of white snow is really easy. Specially compared to the Amazon rainforest, where thick vegetation, and water courses make it impossible."}
library(ggforce)
g +  
  geom_path(data = Compass(0, -75, 180, 15, n = 100), aes(x, y), color = "#ffa154") + 
  annotate("label", x = 0, y = -55, label = "Easy to find") +
  geom_path(data = Compass(-60, -10, 20, 20, n = 100), aes(x, y), color = "#ffa154") +
  annotate("label", x = -100, y = -10, label = "Hard to find") +
  geom_path(data = Compass(-100, 40, 30, 20, n = 100), aes(x, y), color = "#ffa154") +
  annotate("label", x = -160, y = 40, label = "Lots of peple\nlooking?") 
```


```{r, tw = "There are also a lot of badly coded coordinates (0, 0). Either that, or there were a lot of meteorites found on the middle of the Ocean, floating? in the water."}
g + geom_path(data = Compass(0, 0, 5, 5, n = 100), aes(x, y), color = "#ffa154") +
  annotate("label", x = -10, y = -20, label = "Bad\ncoords") 
```
```{r}
meteorites <- meteorites[lat != 0 & long != 0]
```

Much better. Althouth the distribution of falling meteorites is probably uniform, there's clearly a strong bias towards populated areas.


Ok, up untill now it's been kind of still. Let's see some movement. First, has there been an overall tren in meteorite collection? 


```{r, tw = "Ok, let's see some movement. First, has there been an overall tren in meteorite findings? Seems like it. Both types of finds had been increasing steadily, but in the 60's we started to \"find\" a lot of them!"}

(g <- meteorites %>% 
   .[year %between% c(1800, 2013)] %>% 
   .[, .N, by = .(year, fall)] %>% 
   ggplot(aes(year, N)) +
   geom_line() +
   scale_x_continuous(limits = c(NA, 2013), breaks = scales::pretty_breaks(10)) +
   scale_y_log10("N (log)") +
   facet_wrap(~fall, scales = "free_y", ncol = 1) +
   hrbrthemes::theme_ipsum_rc())
```

There are clear trends here. It looks like the anual number of "fell" meteorites had been steadily incrasing from the 1800's until it peaked around the 1940's. From that decade forward, it's been on a sofmore slump, or even slightlly decreasing. Why would that be? Maybe it has to do with the reduction of dark skies that makes it more difficult to actually see anything falling from the heavens. 


```{r}
this_thread$add_post("What do these categories mean, though? A little google lead me to this.", media = "falls_finds.png")
```

```{r, tw = 'Now looking at the ones actually followed from the sky to the ground, it looks like the rate of collection peaked in around the 30s. Why? If I had to guess, I\'d say that it could be related to the lost of dark skies.' }
meteorites %>% 
  .[fall == "Fell"] %>% 
   .[year %between% c(1800, 2013)] %>% 
   .[, .N, by = .(year, fall)] %>% 
   ggplot(aes(year, N)) +
   geom_line() +
   scale_x_continuous(limits = c(NA, 2013), breaks = scales::pretty_breaks(10)) +
   facet_wrap(~fall, scales = "free_y", ncol = 1) +
   hrbrthemes::theme_ipsum_rc() 
```


Yup. the scale is not working here. I'll concentrate on the "fell" meteorites. Bur first, what do those categories mean? According to [this site](https://www.permanent.com/meteorites-falls-finds.html) "fell" meteorites are the ones that have been seen falling from the skies, while "found" meteorites are just found there on the ground. 


```{r}
this_thread$add_post("Now I'm curious about the spacial distribution of these trends. It's a nice challenge. I need to count yearly finds that are near each other to get a representative time series to which to compute a trend.")$
  add_post('The "easy" way would be to lay out a regular grid, but I\'m thinking I want a better option: Hexagon grid. ')
```

Yep, it certrainly looks like a reduction, but those low values after 2010 look fishy to me. Maybe they didn't record them all before the data was released. In any case, I'm also curious about the spatial distribution of trends. 

For that I will divide the globe in hexagons and aggregate the number of finds for each one and each year, and finally compute linear trends. The trends vary wildy, so I use a symmetric log transform on the colours. 

```{r, tw = paste0("I'll use the hexbin package to partition the data into hexagons, aggregate by year and then compute a trend. The result is..",  emo::ji("thinking"), "... not that interesting.")}
library(hexbin)
hex_grid <- meteorites %>% 
  .[fall == "Found"] %>% 
  .[year %between% c(1900, 2010)] %>% 
  with(., hexbin(long, lat, xbins = 15, IDs = TRUE))
hex_grid_df <- as.data.table(hcell2xy(hex_grid)) %>% 
  .[, id := hex_grid@cell]

trends <- meteorites %>% 
  .[fall == "Found"] %>% 
  .[year %between% c(1900, 2010)] %>% 
  .[, id := hex_grid@cID] %>% 
  hex_grid_df[., on = "id"] %>% 
  
  
  .[, .N, by = .(fall, x, y, year)] %>% 
  # .[complete.cases(.)] %>% 
  .[, metR::FitLm(N, year), by = .(fall, x, y)] %>% 
  .[term == "year"] 

C <- 1/log(10)

(g <- trends %>% 
  .[fall == "Found"] %>%
  .[complete.cases(.)] %>%
  .[y > -60] %>%
  ggplot(aes(x, y)) +
  map +
  geom_point(aes(fill = sign(estimate) * log10(1 + abs(estimate/C))), 
             size = 4, shape = 21) +
  metR::scale_fill_divergent("Trends in meteorite finds\n 1900 to 2010 (symmetric log)",
                              oob = scales::squish) +
  metR:::theme_field() )
```

```{r, tw = "Antarctica has seen huge trends, most likely related to the increase in occupation there. But there seems to be an increase also in North Africa, central Australia and in the Atacama desert. "}
g
```

```{r}
this_thread$add_post("Interestingly, it seems that there has ben an increse in north african dealers that sell meteorites to researchers!   http://news.bbc.co.uk/2/hi/africa/6549197.stm")$
  add_post("So this #tidytuesday data provided some in where meteorites are found, opened up the posibility that losing dark skies means less meteorite findings and gave a glimpse into geographic trends.")$
  add_post(paste0("Thread created with the spindler package: https://git.io/fjzxN \n",
                  "It's source code can be found at ", url, "\n",
                  "#rstats"))
```

```{r}
print(this_thread)
```

